Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex

@@

>>>>

\TSComment{we should mention that non-cooperative kernels must be
written in a persistent thread style}

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the mega-kernel), must (a) track the state of all
workgroups, i.e.\ if a workgroup is waiting or computing a kernel (b)
maintain consistent context states for each kernel, i.e.\ the number
of active workgroups computing a cooperative kernel and (c) provide a
safe way for these states to be modified (according to the semantics
of $\offerfork$ and $\offerkill$. We discuss how that is achieved
here.

\paragraph{Kernel context}

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
(starting from zero up to the number of executing workgroups) and the
number of workgroups executing the kernel. In single kernel execution,
the OpenCL primitives {\tt get\_num\_groups()} and {\tt
  get\_group\_id()} provide this functionality natively. For multiple
kernel execution, we must emulate these primitives, per kernel,
explicitly through \emph{kernel contexts}. This structure contains:
(a) the number of active groups, (b) the total number of groups asked
for, and (c) a mapping from the native workgroup id to a new workgroup
id valid for a particular kernel. Our framework requires the native
kernel context (to obtain a global base workgroup id) as well as an
additional distinct kernel context for each kernel running
concurrently. \TSComment{a driver implementation would likely generate
these on the fly}.

\paragraph{Scheduler context}

In order to dynamically manage workgroups executing cooperative
kernels, our framework must (a) track the state of each workgroup and
(b) provide a channel of communication from the scheduler to
workgroups executing $\offerfork$ and $\offerkill$. To achieve this,
we use a \emph{Scheduler context} structure which contains a mapping
from primitive workgroup ids to their state i.e.\ waiting or computing
a kernel as well as an integer value describing the number of
workgroups in each state. The scheduler can then send cooperative
kernels a \emph{resource message}, indicating if workgroups executing
the cooperative kernel need to (a) exit at $\offerkill$ or (b)
transmit their state to forking workgroups at $\offerfork$. Thus, the
scheduler context needs a communication channel for each cooperative
kernel. In our case, the communication channel is implemented using
global atomic variables and OpenCL atomic operations to ensure
consistency.

\paragraph{Launching kernels}

As in native OpenCL, the host maintains the ability to launch kernels.
In our framework, the host has the additional ability to request that
a kernel be cooperative \emph{and} to specify the amount of
\emph{resource} each kernel requires. In our implementation, the only
resource we have explicit control over is the number of concurrently
executing workgroups; implementations closer to metal might also
consider local memory and registers as resource.

To launch a kernel, the host sends a data packet to the GPU, which
consists of: a kernel to execute, the inputs to the kernel, the amount
of resource to use for the kernel (which includes the number of
workgroups to execute the kernel), and a flag indicating if the kernel
is cooperative. In our implementation, this communication channel
between host and device is achieved through fine-grained SVM atomics.
The kernel specified must be one of the kernels merged inside the
mega-kernel. If there are large constant inputs to the kernel, these
can be copied before mega-kernel launch to decrease the amount of SVM
traffic.

The GPU scheduler then receives the data packet from the host
describing a kernel $K$ to be executed and now must make decisions
about how to schedule the $K$. Consider the case where $K$ requires
$W$ workgroups. The scheduler can query the scheduler context and if
there is at least $W$ available workgroups, then $K$ can be scheduled
immediately. If there are not enough available workgroups (say there
are $W_a$ workgroups available), but a cooperative kernel $K_c$ is
executing. The scheduler can signal $K_c$ through $K_c$'s
communication channel in the scheduler context communication
requesting that $K_c$ provide $W - W_a$ workgroups (via $\offerkill$)
as soon as possible. The scheduler then waits until $W$ workgroups
become available.

When $W$ workgroups become available, the scheduler initialises a
kernel context $KC$ for $K$, setting the number of active workgroups
to $W$ and mapping available workgroups (located in the scheduler
context) to their new $KC$ workgroup ids. The scheduler then sends $W$
workgroups from the available workgroups to execute kernel $K$ with
kernel context $KC$.

\paragraph{$\offerkill$ implementation}
